---
title: "ps05-rmarkdown-plot"
author: "Sierra Foster"
date: "2023-02-18"
output: html_document
---
<br>

### 1 Load and check data (5pt)   
**1. (1pt) For solving the problems, and answering the questions, create a new rmarkdown document with an appropriate title.**

```{r, message=FALSE}
library(tidyverse)
```
<br>
<br>
**2. (2pt) Load data. How many rows/columns do we have?**

```{r, message=FALSE}
gap <- read_delim("gapminder.csv.bz2")

cat("We have", nrow(gap), "rows and", ncol(gap), "columns in the dataset.")
```
<br>
<br>
**3. (2pt) Print a small sample of data. Does it look OK?**

Yes, the sample of the data looks okay. It shows each of the column names and prints
out a sample of 5 rows of the data.

```{r}
sample_n(gap, 5)
```
<br>

### 2 Descriptive statistics (15pt)  
**1. (3pt) How many countries are there in the dataset? Analyze all three: iso3, iso2 and name.**

```{r}
iso2uni <- n_distinct(gap$iso2)
iso3uni <- n_distinct(gap$iso3)
nameuni <- n_distinct(gap$name)

cat("There are", nameuni, "unique country names, but there are", iso2uni, "and", iso3uni, 
    "two and three letter abbreviations for countries, meaning there is a possibility of an error in the data.")
```
<br>
<br>
**2. If you did this correctly, you saw that there are more names than iso-2 codes, and there are even more iso3 -codes. What is going on? Can you find it out?**

**(a) (5pt) Find how many names are there for each iso-2 code. Are there any iso-2 codes that correspond to more than one name? What are these countries?**

Based on the codes below, I can conclude that there are 0 codes that correspond 
to more than one name. Each country has 1 code and there are a total of 250 names
with a unique iso2 code. The reason there is 1 less iso2 code then total number 
of names is because one of the names probably has an "NA" field where there iso2
code is supposed to go and there is a missing value.

```{r}
gap %>% 
  group_by(iso2) %>% 
  summarise(num = n_distinct(name)) %>% 
  arrange(desc(num)) %>% 
  count(iso2) %>% 
  head(5)

gap %>% 
  filter(!is.na(iso2)) %>% 
  group_by(iso2) %>% 
  summarise(num = n_distinct(name)) %>% 
  arrange(desc(num)) %>% 
  filter(num>1)
```
<br>
<br>
**(b) (5pt) Now repeat the same for name and iso3-code. Are there country names that have more than one iso3-code? What are these countries? Hint: two of these entitites are CHANISL and NLD CURACAO.**

There are countries that have more than one iso3-code, and the reason that there 
is more iso3-codes then names is because there is "NA" fields where the country
name is supposed to be. Meaning there is country names missing, but iso3-codes are
present, thus adding more iso3-codes then names because there are missing name 
values in the data set.

```{r}
iso3_code <- gap %>% 
  group_by(name) %>% 
  summarise(num = n_distinct(iso3)) %>% 
  filter(num>1)

cat("There are", iso3_code$num, "countries that have more then one iso3-code. These countries are Chanisl, Nld Curacao, and NA values.")
```
<br>
<br>
**3. (2pt) What is the minimum and maximum year in these data?**

```{r}
gap %>% 
  filter(!is.na(time)) %>% 
  summarise(min = min(time), max = max(time))
```
<br>


### 3 CO2 emissions (30pt)
#### Next, let’s analyze CO2 emissions.
**1. (2pt) How many missing co2 emissions are there for each year? Analyze both missing CO2 and co2_PC. Which years have most missing data?**

There are missing CO2 emissions for almost every year, however, the years that
have the most missing data is 2017, 2018, and 2019 with 217 missing values. The next
few being 1960 and 1961 with 60 missing values.

```{r}
missingco2 <- gap %>%
  group_by(time) %>%
  summarise(missing_CO2_emissions = sum(is.na(co2)),
            missing_CO2_PC = sum(is.na(co2_PC)))

missingco2 %>%
  arrange(desc(missing_CO2_emissions + missing_CO2_PC)) %>% 
  head(5)
```
<br>

**2. (5pt) Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few more countries of your choice. Explain what do you see.**

```{r}
gap_subset <- gap %>%
  filter(name %in% c("United States of America", "China", "India", "Italy", "Japan", "Germany"))

ggplot(gap_subset) +
  geom_line(aes(x = time, y = co2, color = name)) +
  labs(title = "Total CO2 Emissions by Country Over Time",
       x = "Year",
       y = "CO2 Emissions (Million Metric Tons)") +
  theme_bw()

```

This graph shows the growth (or decline) in CO2 emissions for China, Germany, 
India, Italy, Japan, and the U.S. Each line represents the amount of CO2 emissions
that have been emitted, and each different colored line represents a different country.
The biggest things that I notice is that China rapidly jumped into leading CO2 emissions
after 2000. They were already high up there, but recently it has taken a huge leap,
but started to decline again in 2015. Around 2005, the U.S. began a decline in CO2 
emissions. India has started incling on CO2 emissions. The rest of the countries seem to be
very leveled with there CO2 emissions, no crazy jumps or declines, just steady.

<br>


**3. (5pt) Now let’s analyze the CO2 emissions per capita (co2_PC). Make a similar plot of the same countries. What does this figure suggest?**

```{r}
gap_subset2 <- gap %>%
  filter(name %in% c("United States of America", "China", "India", "Italy", "Japan", "Germany"))

ggplot(gap_subset2) +
  geom_line(aes(x = time, y = co2_PC, color = name)) +
  labs(title = "CO2 Emissions per Capita by Country Over Time",
       x = "Year",
       y = "CO2 Emissions per Capita (Metric Tons)") +
  theme_bw()

```

With the same layout as the graph above, this graph shows a steady incline over 
the last 50 years in CO2 emissions per capita for India, China, Italy, and Japan. 
Germany has been declining since 1990. Although in the graph above the U.S. was not 
the largest, they are now a significant amount above each of the countries, 
but have started a decline. We were at our highest around 1975. The comparison of both
graphs and the U.S. suggests that even though we do not have the highest overall CO2
emissions, our emissions per person is one of the largest in the world. This means that
when we hear statistics, if it is not per capita, then the U.S. will not be considered the largest
even though per person we have the largest emissions.

<br>


**4. (6pt) Compute average CO2 emissions per capita across the continents (assume region is the same as continent). Comment what do you see. Note: just compute averages over countries and ignore the fact that countries are of different size. Hint: Americas 2016 should be 4.80.**

```{r}
region_co2 <- gap %>% 
  filter(!is.na(co2_PC)) %>% 
  filter(!is.na(region)) %>% 
  group_by(region, time) %>% 
  summarise(avg_co2_PC = mean(co2_PC))
ggplot(region_co2)+
  geom_line(aes(x = time, y = avg_co2_PC, color = region)) +
  labs(title = "CO2 Emissions per Capita in Each Region",
       x = "Year",
       y = "Average CO2 Emissions per Capita") +
    theme_bw()
```
In this graph, I see one line for each of the five regions. the lines stretch out over the 
years (x axis) and depending on there height, is what there average CO2 emissions
per capita will be for that year. It looks like Americas had a spike up in average
CO2 emissions around 1975, but have since then dropped below Europe and Asia. Every
continent seems to be steady in there average emissions, and as a whole look to be declining.
Africa has been steady since 1960, with barely any change in there emissions per capita. 

<br>


**5. (7pt) Make a barplot where you show the previous results–average CO2 emissions per capita across continents in 1960 and 2016. Hint: it should look something along these lines:**

```{r}
gap %>% 
  filter(time %in% c(1960, 2016), !is.na(region), !is.na(co2_PC)) %>% 
  group_by(time, region) %>% 
  summarise(avg_co2_PC = mean(co2_PC)) %>% 
  ggplot(aes(x = region, y = avg_co2_PC, fill = as.factor(time))) +
  geom_col(position = "dodge") +
  labs(title = "Average CO2 Emissions per Capita for Year and Continent",
       x = "Continent",
       y = "Average CO2 Emissions per Capita") +
  scale_fill_discrete(name = "Year")
```

<br>


**6. Which countries are the three largest, and three smallest CO2 emitters (in terms of CO2 per capita) in 2019 for each continent? (Assume region is continent).**

Because 2019 data is missing, I will take 2016 data instead...

```{r}
continent_top3 <- gap %>%
  filter(time == 2016) %>% 
  group_by(region) %>% 
  arrange(desc(co2_PC)) %>% 
  slice_head(n = 3) %>% 
  select(name, region, co2_PC) %>% 
  head(3)

continent_top3
```

```{r}
continent_bottom3 <- gap %>% 
  filter(time == 2016)%>% 
  group_by(region) %>% 
  arrange(co2_PC) %>% 
  slice_head(n = 3) %>% 
  select(name, region, co2_PC) %>% 
  head(3)

continent_bottom3
```
<br>


### 4 GDP per capita (50pt)
#### Let’s look at GDP per capita (GDP_PC).
1. (8pt) Make a scatter plot of GDP per capita versus life expectancy by country, 
using data for 1960. Make the point size dependent on the country size, and color those according to the
continent. Feel free to adjust the plot in other ways to make it better.
Comment what do you see there.

```{r}
gap1960 <- gap %>% 
  filter(!is.na(region), time == 1960)

ggplot(gap1960) +
  geom_point(aes(x = GDP_PC, y = lifeExpectancy, color = region, size = totalPopulation)) +
  scale_size_area(max_size = 10)
```

I see on the x axis, GDP per capita, on the y axis, life expectancy, colored dots
for each region, and the size of dot has to do with the population of the country it 
represents. The curve of the dots suggests that the higher the GDP, the higher the life expectancy
and vise versa. It also suggests that Asia is the biggest region, yet stays on the low end
of GDP with a life expectancy differing from country to country. This scatterplot also shows that
in each region there is a tail end(the country with the least everything) and a high end (a country
that is wealthy with a high life expectancy). So even in regions there is poverty and a 
value of life difference.

2. (4pt) Make a similar plot, but this time use 2019 data only.

```{r}
gap2019 <- gap %>% 
  filter(!is.na(region), time == 2019)

ggplot(gap2019) +
  geom_point(aes(x = GDP_PC, y = lifeExpectancy, color = region, size = totalPopulation)) +
  scale_size_area(max_size = 10)
```

I see the the GDP per capita and life expectancy are effected by each other in 1960. It is shown
that the higher the GDP per capita, the higher the life expectancy goes. It seems that population
is not a huge factor in this graph. The biggest dots seem to be where GDP and life expectancy are
low, but there are also medium sized dots where they are both high, along with smaller dots
everywhere. I also can see that Africa has a low GDP and low life expectancy no matter
the country, but all of the other regions have countries spread out across the scatterplot.

3. (6pt) Compare these two plots and comment what do you see. How has world developed
through the last 60 years?

When comparing these two plots I see that the pattern when comparing life expectancy
and GDP is very noticeable. In the 1960 the pattern follow an upward curve, suggesting
that with a high life expectancy, you have a high GDP. In the graph containing every year,
this same pattern and shape exists. It shows that the world has not developed very much from 1960 in
regards to life expectancy being a factor from a country/regions GDP. I also noticed
that in the graph containg all years, most of the larger dots (countries with large population)
were concentrated in the middle, meaning the larger countries have a harder time reaching that high,
but are advanced enough to not drop to the bottom. In the 1960 graph this is not the case, and
the different sized dots are spread around randomly, suggesting that population might not have
had such a big factor in 1960 as it does now. The world has developed more towards a higher life 
expectancy when the GDP is high and there population is on the smaller side. 

4. (6pt) Compute the average life expectancy for each continent in 1960 and 2019. Do the results
fit with what do you see on the figures?
Note: here as average I mean just average over countries, ignore the fact that countries are of
different size.

```{r}
gap %>% 
  filter(!is.na(region)) %>% 
  filter(!is.na(lifeExpectancy), time == c(1960, 2019)) %>% 
  group_by(region, time) %>%
  summarise(life_mean = mean(lifeExpectancy))
```
Yes, the results fit with what I see on the figures. The GDP over all of the years
is consistnatly higher then the GDP in 1960. In the 1960 graph you can see how low
the numbers for life expectancy and GDP per compared to the graph with ever year.
It suggests that 1960 is a time when these factors were at there lowest across the
world.

5. (8pt) Compute the average LE growth from 1960-2019 across the continents. Show the results
in the order of growth. Explain what do you see.
Hint: these data (data in long form) is not the simplest to compute growth. But you may
want to check out the lag() function. And do not forget to group data by continent when
using lag(), otherwise your results will be messed up! See https://faculty.washington.
edu/otoomet/info201-book/dplyr.html#dplyr-helpers-compute.

```{r}
gap %>%
  filter(time >= 1960 & time <= 2019) %>% 
  filter(!is.na(region)) %>% 
  group_by(region) %>%
  mutate(avg_growth = (lifeExpectancy - lag(lifeExpectancy))/lag(lifeExpectancy)) %>%
  summarise(avg_growth = mean(avg_growth, na.rm = TRUE)) %>%
  arrange(desc(avg_growth))
```
From this code I see that Oceania has had the most growth in there life
expectancy from 1960 to 2019. Europe and Americas are at the bottom. This is 
suprising to me because of the idea that we are more developed in the healthcare and
life expectancy area, however it seems that it might not be as true. It might be 
the way people have changed over time (more lazy, unhealthy habits) that has made 
Europe and Americas fall behind in the growth of life expectancy. It could also be
that we are leveling out what technologies we have, and Africa and Oceania have just
gotten access to those technologies thus making there life expectancy grow at a must
faster rate then others due to them catching up to the other 3 continents. Also, even
though Africa had one of the worst GDP to life expectancy ratios, they have had the 
second most growth in life expectancy. 

6. (6pt) Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both
histograms on the same graph, see how well you can do it!

```{r}
gap %>% 
  filter(time %in% c(1960, 2019)) %>%
  ggplot(aes(x = GDP_PC, fill = factor(time))) +
  geom_histogram(position = "dodge", color = "black", binwidth = 5000) +
  scale_fill_manual(values = c("blue", "red"), name = "Year", labels = c("1960", "2019")) +
  labs(x = "GDP per capita", y = "Frequency", title = "Histogram of GDP per capita in 1960 and 2019") +
  theme(plot.title = element_text(hjust = 0.5))
```


7. (6pt) What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When
counting from top.)
Hint: check out the function rank()!
Hint2: 17 for 1960.

```{r}
USrank <- gap %>%
  filter(!is.na(name)) %>% 
  group_by(time) %>%
  mutate(rank = rank(-lifeExpectancy)) %>%
  filter(name == "United States of America" & time %in% c(1960, 2019)) %>%
  select(time, rank)
USrank
```

The rank for the United States in terms of life expectancy was 17 in 1960
and 46 in 2019, meaning other countries have surpassed us since 1960.

8. (6pt) If you did this correctly, then you noticed that US ranking has been falling quite a
bit. But we also have more countries in 2019–what about the relative rank divided by the
corresponding number of countries that have LE data in the corresponding year?
Hint: 0.0904 for 1960.

```{r}

countries <- gap %>%
  filter(!is.na(lifeExpectancy)) %>%
  filter(!is.na(name)) %>% 
  group_by(time) %>%
  summarize(num = n()) %>% 
  filter(time %in% c(1960, 2019))

cat("The rank divided by the number of countries for that year is as follows:\n",
    17/188, "for 1960 and", 46/196, "for 2019 based on number of countries.")
```


#### Finally tell us how many hours did you spend on this PS.

I spent about 10 hours on this assignment. I do a lot of stopping and going
throughout the day.

