---
title: "ps05-rmarkdown-plot"
author: "Sierra Foster"
date: "2023-02-18"
output: html_document
---

### 1 Load and check data (5pt)
#### You first task is to do a very simple data check:
1. (1pt) For solving the problems, and answering the questions, create a new rmarkdown document with an appropriate title. See https://faculty.washington.edu/otoomet/info201-book/
r-markdown.html#r-markdown-rstudio-creating.

```{r, message=FALSE}
library(tidyverse)
```


2. (2pt) Load data. How many rows/columns do we have?

```{r, message=FALSE}
gap <- read_delim("gapminder.csv.bz2")
```

```{r}
nrow(gap)
ncol(gap)
```

3. (2pt) Print a small sample of data. Does it look OK?

Yes, the sample of the data looks okay. It shows each of the column names and prints
out a sample of 5 rows of the data.

```{r}
sample_n(gap, 5)
```

### 2 Descriptive statistics (15pt)
1. (3pt) How many countries are there in the dataset? Analyze all three: iso3, iso2 and name.

```{r}
iso2uni <- n_distinct(gap$iso2)
iso3uni <- n_distinct(gap$iso3)
nameuni <- n_distinct(gap$name)

cat("There are", nameuni, "unique country names, but there are", iso2uni, "and", iso3uni, 
    "two and three letter abbreviations for countries, meaning there is a possibility of an error in the data.")
```

2. If you did this correctly, you saw that there are more names than iso-2 codes, and there are
even more iso3 -codes. What is going on? Can you find it out?

(a) (5pt) Find how many names are there for each iso-2 code. Are there any iso-2 codes that
correspond to more than one name? What are these countries?



(b) (5pt) Now repeat the same for name and iso3-code. Are there country names that have
more than one iso3-code? What are these countries?
Hint: two of these entitites are CHANISL and NLD CURACAO.



3. (2pt) What is the minimum and maximum year in these data?



### 3 CO2 emissions (30pt)
#### Next, let’s analyze CO2 emissions.
1. (2pt) How many missing co2 emissions are there for each year? Analyze both missing CO2
and co2_PC. Which years have most missing data?

```{r}
missingco2 <- gap %>%
  group_by(time) %>%
  summarise(missing_CO2_emissions = sum(is.na(co2)),
            missing_CO2_PC = sum(is.na(co2_PC)))

missingco2 %>%
  arrange(desc(missing_CO2_emissions + missing_CO2_PC)) %>%
  head(10)
```


2. (5pt) Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few
more countries of your choice. Explain what do you see.

```{r}
gap_subset <- gap %>%
  filter(name %in% c("United States of America", "China", "India", "Italy", "Japan", "Germany"))

ggplot(gap_subset) +
  geom_line(aes(x = time, y = co2, color = name)) +
  labs(title = "Total CO2 Emissions by Country Over Time",
       x = "Year",
       y = "CO2 Emissions (Million Metric Tons)") +
  theme_bw()

```

__This graph shows the growth (or decline) in CO2 emissions for China, Germany, 
India, Italy, Japan, and the U.S. Each line represents the amount of CO2 emissions
that have been emitted, and each different colored line represents a different country.
The biggest things that I notice is that China rapidly jumped into leading CO2 emissions
after 2000. They were already high up there, but recently it has taken a huge leap,
but started to decline again in 2015. Around 2005, the U.S. began a decline in CO2 
emissions. India has started incling on CO2 emissions. The rest of the countries seem to be
very leveled with there CO2 emissions, no crazy jumps or declines, just steady.__


3. (5pt) Now let’s analyze the CO2 emissions per capita (co2_PC). Make a similar plot of the
same countries. What does this figure suggest?

```{r}
gap_subset2 <- gap %>%
  filter(name %in% c("United States of America", "China", "India", "Italy", "Japan", "Germany"))

ggplot(gap_subset2) +
  geom_line(aes(x = time, y = co2_PC, color = name)) +
  labs(title = "CO2 Emissions per Capita by Country Over Time",
       x = "Year",
       y = "CO2 Emissions per Capita (Metric Tons)") +
  theme_bw()

```

This graph shows 

4. (6pt) Compute average CO2 emissions per capita across the continents (assume region is the
same as continent). Comment what do you see.
Note: just compute averages over countries and ignore the fact that countries are of different
size.
Hint: Americas 2016 should be 4.80.



5. (7pt) Make a barplot where you show the previous results–average CO2 emissions per capita
across continents in 1960 and 2016.
Hint: it should look something along these lines:



6. Which countries are the three largest, and three smallest CO2 emitters (in terms of CO2 per
capita) in 2019 for each continent? (Assume region is continent).



### 4 GDP per capita (50pt)
#### Let’s look at GDP per capita (GDP_PC).
1. (8pt) Make a scatterplot of GDP per capita versus life expectancy by country, 
using data for 1960. Make the point size dependent on the country size, and color those according to the
continent. Feel free to adjust the plot in other ways to make it better.
Comment what do you see there.



2. (4pt) Make a similar plot, but this time use 2019 data only.



3. (6pt) Compare these two plots and comment what do you see. How has world developed
through the last 60 years?



4. (6pt) Compute the average life expectancy for each continent in 1960 and 2019. Do the results
fit with what do you see on the figures?
Note: here as average I mean just average over countries, ignore the fact that countries are of
different size.



5. (8pt) Compute the average LE growth from 1960-2019 across the continents. Show the results
in the order of growth. Explain what do you see.
Hint: these data (data in long form) is not the simplest to compute growth. But you may
want to check out the lag() function. And do not forget to group data by continent when
using lag(), otherwise your results will be messed up! See https://faculty.washington.
edu/otoomet/info201-book/dplyr.html#dplyr-helpers-compute.



6. (6pt) Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both
histograms on the same graph, see how well you can do it!



7. (6pt) What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When
counting from top.)
Hint: check out the function rank()!
Hint2: 17 for 1960.



8. (6pt) If you did this correctly, then you noticed that US ranking has been falling quite a
bit. But we also have more countries in 2019–what about the relative rank divided by the
corresponding number of countries that have LE data in the corresponding year?
Hint: 0.0904 for 1960.



#### Finally tell us how many hours did you spend on this PS.


